Fail Faster

### 2025-03-16 paper/paper.tex
less rough

\jwc{This paper is about ``Strict functional languages like OCaml and Scala'', but we'll focus on OCaml for presentation.}

\bcp{The current introduction says, roughly, ``We took a look at
  base.quickcheck, noticed some inefficiencies, applied known
  techniques from staged metaprogramming to eliminate them, and
  observed that speed went up.''  I think we can make much stronger
  claims, but I'm not sure about exactly how strong or what belongs in
  the foreground...
  \begin{itemize}
  \item One of the first serious uses of staging in anger?
  \item A use of staging that is novel / challenging in
  itself, in some way?
  \item An analysis of sources of inefficiency across a range of PBT
  frameworks (and a solution that applies to many of them)?
  \item A usable tool that addresses and overcomes some significant
  implementation challenges?
  \item Careful measurements showing where the sources of inefficiency
  are in existing PBT tools, and which ones matter most?
  \item (Your claim here...)?
  \end{itemize}
}

\jwc{Agree: I think something like the 3rd is the best reframing. Something
like: ``We demonstrate that, across languages, monadic PBT generator DSLs can
have a significant performance overhead, and present a cross-language technique
for eliminating the overhead while preserving the idiomatic abstraction.
''}

\jwc{People have been using staged metaprogrammign to eliminate abstraction
overheads in parsing for a long time, we turn that lens on generation (Which looks like parsing)}

\jwc{Another angle that I'd argue needs to come out more is that we demonstrate that
engineering PBT libraries with performance in mind has significant impact on testing power.
}

\jwc{
Let's lead on with the quote from PBT in practice: performance engineering
matters.  De-emphasize the abstraction overhead framing. Eliminating abstraction
overhead is important, so is using the fastest RNG you can possibly use.
}

\jwc{``Property based testing is a race against time...''}

Property-based testing (PBT) uses generators of random data to produce inputs to a
system under test, which is then checked against a set of properties the system is
 expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
 to software correctness, aimed at detecting bugs within a short testing 
 window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
 it is imperative that inputs be generated as fast as possible; in other words, 
 generators must be treated as performance-sensitive code.

 The locus for improvement in this area is \textit{generator libraries}: if programmers are 
 provided with the tools to write fast generators, then they do not themselves 
 need to be performance experts. However, existing libraries fail to achieve good performance 
 due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
 of selecting random values. (Another sentence here?)
 The second is abstraction: generator libraries are implemented as embedded DSLS, which are flexible
 and expressive, but also introduce significant overhead that is not necessarily optimized away by 
 the compiler. These costs compound in the aggregate, raising a fundamental
 question: can generator libraries produce code that is both expressive \textit{and} efficient?

  % in strict functionallanguages like OCaml and Scala
We outline a fully general, two-stage approach to designing generator libraries, 
which provide high-level combinator functions with dramatically reduced 
overhead. First, \rand choice matters. We demonstrate that using a faster \rand
significantly impacts generator performance by directly optimizing an existing RNG, which
produces the same values as its unoptimized counterpart---but faster--thereby allowing us
to compare the RNG's impact on an existing generator library. Second, we 
describe a method of optimizing generator libraries using \textit{staged metaprogramming}, 
or ``staging''. Staged metaprogramming allows for code generation, which can mitigate 
performance challenges introduced by embedded DSLs. In particular, when a generator DSL 
introduces overhead for dynamic control flow, staging can specialize that control flow
and compile it statically. For example, when choosing between a list of options, a staged
generator can avoid allocating the list of options at runtime by expanding to an 
if-else. This approach is not tied to any particular PBT library---instead, it gives a recipe
for improving even SOTA PBT libraries.

To demonstrate
the effectiveness of our technique, we apply our technique to a leading PBT library, \bq, built with performance engineering in mind.
Our version of this library, termed \name, is a direct drop-in replacement for the original that is completely
semantically equivalent: generators written using it produce exactly the same values as 
those written in the original version. However, in \name, the abstractions provided by the generator DSL are 
zero-cost, and the \rand is highly-optimized. The direct equivalence between \name and \bq
allows us to perform extremely fine-grained comparisons between generators---generators implemented with both versions of the
library will produce exactly the same sequence of values, so differences in testing performance are directly attributable
to \name's performance speedup.

To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
\bq on a variety of different workloads. We demonstrate the generality of the technique by implementing \name in
Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate
that \name allows you to find bugs significantly faster ...

We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter concern based on staged metaprogramming. \secref{section:motiv}
    \item We implement a version of \bq, \name, that contains our optimizations and maintains strict program equality. \secref{section:4}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}

>>>>

%\jwc{This paper is about ``Strict functional languages like OCaml and Scala'', but we'll focus on OCaml for presentation.}

%\bcp{The current introduction says, roughly, ``We took a look at
%  base.quickcheck, noticed some inefficiencies, applied known
%  techniques from staged metaprogramming to eliminate them, and
%  observed that speed went up.''  I think we can make much stronger
%  claims, but I'm not sure about exactly how strong or what belongs in
%  the foreground...
%  \begin{itemize}
%  \item One of the first serious uses of staging in anger?
%  \item A use of staging that is novel / challenging in
%  itself, in some way?
%  \item An analysis of sources of inefficiency across a range of PBT
%  frameworks (and a solution that applies to many of them)?
%  \item A usable tool that addresses and overcomes some significant
%  implementation challenges?
%  \item Careful measurements showing where the sources of inefficiency
%  are in existing PBT tools, and which ones matter most?
%  \item (Your claim here...)?
%  \end{itemize}
%}

%\jwc{Agree: I think something like the 3rd is the best reframing. Something
%like: ``We demonstrate that, across languages, monadic PBT generator DSLs can
%have a significant performance overhead, and present a cross-language technique
%for eliminating the overhead while preserving the idiomatic abstraction.
%''}

%\jwc{People have been using staged metaprogrammign to eliminate abstraction
%overheads in parsing for a long time, we turn that lens on generation (Which looks like parsing)}

%\jwc{Another angle that I'd argue needs to come out more is that we demonstrate that
%engineering PBT libraries with performance in mind has significant impact on testing power.
%}

%\jwc{
%Let's lead on with the quote from PBT in practice: performance engineering
%matters.  De-emphasize the abstraction overhead framing. Eliminating abstraction
%overhead is important, so is using the fastest RNG you can possibly use.
%}

%\jwc{``Property based testing is a race against time...''}

Property-based testing (PBT) is a testing technique where random data
produced by a \textit{generator} is fed
to a system under test, which is then checked against a set of properties 
it is expected to satisfy. Recent work on understanding how PBT is used in 
practice has determined that practitioners often run PBT with very short time 
budgets---typically 50 milliseconds to 30 seconds. Given this practical 
constraint, it is imperative that inputs be generated as fast as possible; 
in other words, generators must be treated as performance-sensitive code.
A high-leverage way to improve the performance of generators is to optimize 
\textit{generator libraries}: if
programmers are provided with the tools to write fast generators, then they 
do not themselves need to be performance experts.

However, the performance of 
existing generator libraries falls short of what is possible due to two major 
sources of inefficiency. The first of these is \textit{abstraction}: while 
flexible and expressive, generator libraries rely on complex abstractions
and indirections that are difficult for compilers to fully simplify, resulting
 in additional function calls and closure allocations.
Our second insight involves the process of 
selecting random values (``random sampling''), which, depending on one's choice of \rand and
the number of samples needed, can constitute an unexpectedly high proportion
of a generator's runtime. This raises a fundamental 
question: can generator libraries produce code that is both expressive 
\textit{and} efficient?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

% The locus for improvement in this area \jwc{``This area'' sounds like it's talking about the whole area of PBT... Maybe ``a high-leverage way to improve the performance of generators is to...''} is \textit{generator libraries}: if

  % in strict functionallanguages like OCaml and Scal
To answer this, we propose a two-pronged approach to designing generator libraries 
that preserves their expressiveness while eliminating unnecessary overhead.
First, we propose a method of optimizing
generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
long been used across domains as a technique of eliminating the overhead 
of idiomatic functional DSLs \todo{CITE!!!}.
Using staging, it is possible to eliminate the overhead of many common 
generator abstractions. For example,
each call to monadic bind in \bq constructs and then immediately executes a 
closure at runtime; meanwhile, in \name, bind resolves at compile time,
allowing the compiler to inline away the closures. Secondly, we demonstrate the
impact of different choices of \rand on generator performance 
by conducting a controlled comparison. Together, these insights provide a
recipe for improving even state-of-the-art generator libraries.

%\jwc{I think we should use return/bind as the simplest example, since then you don't have to get into the weeds of allocation or closures. The functional abstractions of a monadic generator DSL block the compiler from generating fast code.}
%\jwc{Then present this case as ``more sophisticated use of staging'' or something?}
%where an unstaged generator DSL might represent a choice between
%a list of options as a dynamically-allocated array of closures, a staged 
%generator can expand to an if-else tree at compile 
%time---eliminating the allocation entirely.
% This approach is not tied to any
%particular PBT library, but rather provides a recipe for improving even
%state-of-the-art implementations \jwc{Sentence structure doesn't follow. Instead maybe ``not any particular, but a recipe for improving in general''}.

%First, we demonstrate the impact of
%RNG choice by optimizing an existing \rand. Our optimized RNG produces 
%exactly 
%the same values as its unoptimized counterpart, allowing us to directly 
%measure the impact of this choice on generator performance. \jwc{Permute? I think staging \emph{then} randomness.} \jwc{Also, i don't love this as the first place we explain the drop-in equivalence of the RNG. It's unmotivated why wen want this.}

To demonstrate the effectiveness of our technique, we apply it to \bq,
which is the fastest property-based testing library in OCaml according to our tests.
%\jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
Our version of this library, \name, is a drop-in replacement for \bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name 
produce exactly the same values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons between generators, as generators 
implemented using both versions of the library will produce exactly the same 
sequence of values. Therefore, differences in performance are directly 
attributable to \name's speedup.

To evaluate \name's effectiveness, 
implement a series of workloads in each generator library, and demonstrate that 
\name runs faster and allocates less than \bq. Further, we show that this performance benefit
translates to bug-finding ability by running these workloads in Etna, a PBT platform that 
simulates real-world usage by injecting bugs into a system and measuring how quickly different 
generators can detect them.
Finally, to demonstrate the generality of our technique, we implement \name in Scala
and benchmark the same generators in the Scala port of the library---thereby highlighting 
its portability
to other strict functional languages and reinforcing its potential for widespread adoption.

In summary, we demonstrate that, across languages, monadic PBT generator DSLs
have a significant performance overhead, and present a cross-language technique
for eliminating that overhead while preserving their idiomatic abstractions.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter based on staged metaprogramming. \secref{section:motiv}
    \item We implement a version of \bq, \name, that showcases our optimization technique while maintaining strict program equality. \secref{section:4}
    \item We evaluate \name by demonstrating the improved performance of generators constructed
    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
\end{enumerate}
